Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2023-06-21 manuscripts/main.tex
manuscript proofs

% load packages
\usepackage[utf8]{inputenc}
\usepackage[a4paper, margin=0.8in]{geometry}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{hyperref}
\usepackage{float}
\usepackage{amsmath}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[flushmargin]{footmisc}
\usepackage[flushleft]{threeparttable}
\usepackage{array}
\usepackage[font=small, labelfont=it]{caption}
\usepackage{booktabs}
\usepackage{xcolor,soul}

@@

% define highlight style
\colorlet{usercolorname}{yellow!00}    % use for public version
% \colorlet{usercolorname}{yellow!30}    % use for reviewer version
\sethlcolor{usercolorname}

@@

\maketitle

@@

Finally, we compared the proportion of participants meeting the cutoff for clinical levels of psychopathology before and after excluding participants based on their responses to the infrequency items. Previous studies have found that applying such measures reduced the prevalence of clinical symptomology in online samples towards ground truth rates from epidemiological studies \cite{ophir2020turker}. On the most positively-skewed measures, the fraction of participants reaching clinical levels of symptom endorsement prior to screening was greater than what would be expected (Table \ref{tab:tab02}). For example, 13.0\% of participants scored at or above clinical thresholds for (hypo)mania on the 7-up scale in our sample prior to screening, compared with a 12-month prevalence of 5\% in the general population \cite{merikangas2007lifetime, merikangas2012true}, but this rate was reduced to 4.0\% (in line with the population prevalence estimates) after exclusion of C/IE respondents. We observed a similar pattern for both major depressive disorder and anxiety (population prevalence estimates of 7\% and 5\% respectively; \cite{kessler2012twelve, lowe2008validation, hinz2017psychometric}). Interestingly, the proportion of participants meeting threshold on the GAD-7 was elevated compared to previous literature. We suspect this may reflect elevated rates of state anxiety during the COVID-19 pandemic \cite{yarrington2021impact}, when these data were collected. In line with previous research, we interpret these inflated rates of clinical symptomology in our sample prior to screening as suggestive of C/IE responding \cite{ophir2020turker}.

@@

To evaluate the generalizability of our findings, we therefore conducted a conceptual replication experiment in which an independent sample of N=393 participants (N=193 from MTurk using CloudResearch, N=200 from Prolific) completed a more difficult cognitive task, the well-known ``two-step task'' \cite{daw2011model}, and an alternate set of self-report measures (see Appendix B for details). Importantly, participants were recruited \textit{after} CloudResearch and Prolific implemented new protocols to improve data quality on their respective platforms. As a final control measure, participants completed self-report symptom measures as before, but also personality measures with no hypothesized relationship to model-based planning behavior on the two-step task.

For the sake of brevity, we report here only the main pattern of findings (all results are reported in Appendix B). In the replication sample, 55 out of 393 participants (14\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report measures. This is roughly two-thirds of the fraction of participants who were flagged for C/IE responding in the original study, suggesting that the newer quality assurance protocols used by the online platforms are at least partially effective.

@@

% \subsection*{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}
\subsection*{Psychiatric patients do not fail attention checks more often}

@@

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Appendix C for details). Specifically, N=35 psychiatric patients (confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview) across 45 unique testing sessions and N=17 healthy controls across 20 unique testing sessions, all recruited through the Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry (i.e., not via online labor platforms), completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) psychiatric participants were flagged for C/IE responding.

Using these data, we computed pairwise Bayes factors comparing three candidate models: attention check failure rates are equal between healthy and psychiatric participants (M1); failure rates are greater in psychiatric patients (M2); and failure rates are greater in healthy participants (M3). The model assuming equal rates of failure between healthy and psychiatric participants was 2.88 times more likely than the model assuming greater rates for psychiatric patients. In turn, the model assuming lower rates of failure for psychiatric patients was 1.27 times more likely than the model assuming equal rates. Finally, the model assuming lower rates of failure for psychiatric patients was 3.65 times more likely than the model assuming higher rates for psychiatric patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a difference in model fit as meaningful. Although the size of the sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where psychiatric patients are more likely to fail infrequency item attention checks. These data suggest, therefore, that it is unlikely that individuals with high psychiatric symptom severity were disproportionately flagged for C/IE responding in the main analyses. Accordingly, we tentatively conclude that the screening measures we are suggesting are not likely to result in overcontrol bias and false-negative correlations between tasks and symptom measures.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g., low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence that depressed patients were more likely to fail attention checks than healthy controls (if anything, non-psychiatric participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias. However, further research with larger samples is necessary to validate attention checks in depressed and other patient populations.

@@

When collecting self-report questionnaire data, we encourage experimenters to use screening methods sensitive to multiple distinct patterns of C/IE responding (e.g., random responding, straight-lining, side bias) and, if possible, to log all page interactions (e.g., mouse clicks, keyboard presses). We specifically recommend experimenters use infrequency-item attention checks rather than instructed-item checks, as multiple studies have now shown that online participants are habituated to and circumvent the latter (e.g., \cite{barends2019noncompliant, thomas2017validity, hauser2016attentive}; Appendix B). Participants flagged by suspicious responses on attention-check items should either be excluded from further analysis, or assessed using sensitivity analyses to ensure that observed full-sample correlations are robust to their exclusion.

@@

Data from several participants were excluded prior to analysis. Three participants (all MTurk) were excluded due to missing data. In addition, we excluded 20 participants who disclosed that they had also completed the experiment on the other platform. This left a final sample of N=386 participants (MTurk: N=186, Prolific: N=200) for analysis. The demographics of the sample split by labor market is provided in Table S1. Notably, the participants recruited from MTurk were older ($M = 7.7 \ \text{yrs}$, two-tailed, two-sample $t$-test: $t(384)=6.567$, $p<0.001$, $d=0.669$, 95\% CI = [5.4, 10.0]) and comprised of fewer women (two-tailed, two-sample proportions test: $z(384)=2.529$, $p=0.011$, $h=0.258$, 95\% CI = [0.030, 0.228]).

@@

As a cover story, the probabilistic reversal learning task was introduced to participants as a fishing game in which each choice option was a beach scene made distinguishable by a colored surfboard with unique symbol. Participants were told they were choosing which beach to fish at. Feedback was presented as either a fish (1 point) or trash (0 points). Participants were instructed to earn the most points possible by learning (through trial-and-error) and choosing the best choice option. Participants were also instructed that the best option could change during the task, but were not informed about how often or when this would occur (see Appendix A for the complete instructions). Prior to beginning the experiment, participants had to correctly answer four comprehension questions about the instructions. Failing to correctly answer all items forced the participant to start the instructions over.

@@

\subparagraph{Clinical cutoffs.} We compared the proportion of participants in our sample reaching the threshold for clinical symptomology before and after applying exclusions. For the GAD-7, previous research has suggested a clinical cutoff score of 10 or higher \cite{lowe2008validation, hinz2017psychometric}. Though the 7-up/7-down scales do not have firmly established clinical cutoffs recent work has suggested a cutoff score of 12 or higher \cite{youngstrom2020evaluating}, which we use here. Finally, the original authors of the SHAPS scale recommended as a cutoff a score of 3 or more when the items are binarized (1 = ``strongly disagree", ``agree"; 0 = ``strongly agree", ``agree"). We use this scoring approach in Table \ref{tab:tab02}.

@@

where $y_i$ is a participant's response to item $i$, $x_i$ is a participant's average score across all  items, and $k$ is the total number of items for a self-report measure. A composite ISD measure was estimated per participant by summing across each of the seven self-report scales. Larger ISD values indicate lower response consistency.

@@

In the risk-sensitive temporal difference learning model, there are separate learning rates for positive and negative prediction errors, such that positive and negative prediction errors have asymmetric effects on learning. For example, the effect of negative prediction errors on learned values is larger than that of positive errors if $\eta_p+ < \eta_n$, and vice versa if $\eta_p > \eta_n$.

@@

\textbf{Samuel Zorowitz:} Conceptualization (equal); Software development (lead); Data collection - online (lead); Formal analysis (lead); Writing - Original Draft (lead); Writing - Review \& Editing (supporting); Visualization (lead). \textbf{Johanne Solis:} Software development (supporting); Data collection - clinical (lead); Writing - Review \& Editing (supporting). \textbf{Yael Niv:} Writing - Review \& Editing (equal); Funding acquisition. \textbf{Daniel Bennett:} Conceptualization (equal); Software development (supporting); Data collection - online (supporting); Formal analysis (supporting); Writing - Review \& Editing (equal); Visualization (supporting).

>>>>

% load packages
\usepackage[utf8]{inputenc}
\usepackage[a4paper, margin=0.8in]{geometry}
\usepackage{authblk}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{indentfirst}
\usepackage{hyperref}
\usepackage{float}
\usepackage{amsmath}
\usepackage[backend=biber,style=nature]{biblatex}
\usepackage[flushmargin]{footmisc}
\usepackage[flushleft]{threeparttable}
\usepackage{array}
\usepackage[font=small, labelfont=it]{caption}
\usepackage{booktabs}
\usepackage{fancyhdr}           % first-page header

@@

% specify first-page header
\fancypagestyle{firstpage}
{
    \renewcommand{\headrulewidth}{0pt}
    \fancyhead[L]{\small Peer-reviewed publication DOI: \href{https://doi.org/10.1038/s41562-023-01640-7}{10.1038/s41562-023-01640-7}}
    \fancyhead[R]{}
}

@@

\maketitle
\thispagestyle{firstpage}

@@

Finally, we compared the proportion of participants meeting the cutoff for clinical levels of psychopathology before and after excluding participants based on their responses to the infrequency items. Previous studies have found that applying such measures reduced the prevalence of clinical symptomology in online samples towards ground truth rates from epidemiological studies \cite{ophir2020turker}. On the most positively-skewed measures, the fraction of participants reaching clinical levels of symptom endorsement prior to screening was greater than what would be expected (Table \ref{tab:tab02}). For example, 13.0\% of participants scored at or above clinical thresholds for (hypo)mania on the 7-up scale in our sample prior to screening, compared with a 12-month prevalence of 5\% in the general population \cite{merikangas2007lifetime, merikangas2012true}, but this rate was reduced to 4.0\% (in line with the population prevalence estimates) after exclusion of C/IE respondents. We observed a similar pattern for both major depressive disorder (MDD) and anxiety (population prevalence estimates of 7\% and 5\% respectively; \cite{kessler2012twelve, lowe2008validation, hinz2017psychometric}). Interestingly, the proportion of participants meeting threshold on the GAD-7 was elevated compared to previous literature. We suspect this may reflect elevated rates of state anxiety during the COVID-19 pandemic \cite{yarrington2021impact}, when these data were collected. In line with previous research, we interpret these inflated rates of clinical symptomology in our sample prior to screening as suggestive of C/IE responding \cite{ophir2020turker}.

@@

To evaluate the generalizability of our findings, we therefore conducted a conceptual replication experiment in which an independent sample of N=393 participants (N=193 from MTurk using CloudResearch, N=200 from Prolific) completed a more difficult cognitive task, the well-known ``two-step task'' \cite{daw2011model}, and an alternate set of self-report measures (see Supplementary Materials B for details). Importantly, participants were recruited \textit{after} CloudResearch and Prolific implemented new protocols to improve data quality on their respective platforms. As a final control measure, participants completed self-report symptom measures as before, but also personality measures with no hypothesized relationship to model-based planning behavior on the two-step task.

For the sake of brevity, we report here only the main pattern of findings (all results are reported in Supplementary Materials B). In the replication sample, 55 out of 393 participants (14\%) endorsed a logically invalid or improbable response on one or more of the infrequency items when completing the self-report measures. This is roughly two-thirds of the fraction of participants who were flagged for C/IE responding in the original study, suggesting that the newer quality assurance protocols used by the online platforms are at least partially effective.

@@

% \subsection*{Individuals with and without diagnosed psychiatric disorders fail attention checks at similar rates}
\subsection*{Patients with depression do not fail attention checks more often}

@@

To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder (see Supplementary Materials C for details). Specifically, N=35 psychiatric patients (confirmed to meet criteria for a diagnosis of major depressive disorder through a structured clinical interview) across 45 unique testing sessions and N=17 healthy controls across 20 unique testing sessions, all recruited through the Rutgers-Princeton Center for Computational Cognitive Neuropsychiatry (i.e., not via online labor platforms), completed a series of self-report symptom measures, online, on their computers from the comfort of their homes. In total, 16 of 65 (24.6\%) participants failed one or more attention checks. Subdivided by group, 6 of 20 (30\%) healthy participants and 10 of 45 (22\%) MDD patients were flagged for C/IE responding.

Using these data, we computed pairwise Bayes factors comparing three candidate models: attention check failure rates are equal between healthy and MDD patients (M1); failure rates are greater in MDD patients (M2); and failure rates are greater in healthy participants (M3). The model assuming equal rates of failure between healthy and MDD participants was 2.88 times more likely than the model assuming greater rates for MDD patients. In turn, the model assuming lower rates of failure for MDD patients was 1.27 times more likely than the model assuming equal rates. Finally, the model assuming lower rates of failure for MDD patients was 3.65 times more likely than the model assuming higher rates for MDD patients. Only the final comparison exceeds the cutoff value of 3, which is conventionally treated as the minimal amount of evidence required to treat a difference in model fit as meaningful. Although the size of the sample precludes any definitive conclusion, it is noteworthy that the model least consistent with the data was the one where MDD patients are more likely to fail infrequency item attention checks. These data suggest, therefore, that it is unlikely that individuals with high depression symptom severity were disproportionately flagged for C/IE responding in the main analyses. Accordingly, we tentatively conclude that the screening measures we are suggesting are not likely to result in overcontrol bias and false-negative correlations between tasks and symptom measures, at least in the case of individuals with depression. It remains possible that other psychiatric symptoms might be associated with a different pattern of results.

@@

One objection to the rigorous screening and exclusion of participants based on C/IE detection methods is that we might inadvertently introduce an overcontrol bias. That is, to the extent that C/IE responding might reflect symptoms common to psychopathology (e.g., low motivation, effort avoidance, inattentiveness), rigorous screening of C/IE responding might lead to the differential exclusion of truly symptomatic participants. To explore this possibility, we embedded attention checks into the self-report measures of two studies of patients with major depressive disorder. Though our final sample was small, we did not find evidence that depressed patients were more likely to fail attention checks than healthy controls (if anything, healthy participants were more likely to be flagged by C/IE screening). These results provide preliminary evidence that rigorous C/IE screening is unlikely to result in overcontrol bias. However, further research with larger samples is necessary to validate attention checks in depressed and other patient populations.

@@

When collecting self-report questionnaire data, we encourage experimenters to use screening methods sensitive to multiple distinct patterns of C/IE responding (e.g., random responding, straight-lining, side bias) and, if possible, to log all page interactions (e.g., mouse clicks, keyboard presses). We specifically recommend experimenters use infrequency-item attention checks rather than instructed-item checks, as multiple studies have now shown that online participants are habituated to and circumvent the latter (e.g., \cite{barends2019noncompliant, thomas2017validity, hauser2016attentive}; Supplementary Materials B). Participants flagged by suspicious responses on attention-check items should either be excluded from further analysis, or assessed using sensitivity analyses to ensure that observed full-sample correlations are robust to their exclusion.

@@

Data from several participants were excluded prior to analysis. Three participants (all MTurk) were excluded due to missing data. In addition, we excluded 20 participants who disclosed that they had also completed the experiment on the other platform. This left a final sample of N=386 participants (MTurk: N=186, Prolific: N=200) for analysis. The demographics of the sample split by labor market is provided in Table S1. Notably, the participants recruited from MTurk were older (mean difference = 7.7 yrs), two-tailed, two-sample $t$-test: $t(384)=6.567$, $p<0.001$, $d=0.669$, 95\% CI = [5.4, 10.0]) and comprised of fewer women (two-tailed, two-sample proportions test: $z(384)=2.529$, $p=0.011$, $h=0.258$, 95\% CI = [0.030, 0.228]).

@@

As a cover story, the probabilistic reversal learning task was introduced to participants as a fishing game in which each choice option was a beach scene made distinguishable by a colored surfboard with unique symbol. Participants were told they were choosing which beach to fish at. Feedback was presented as either a fish (1 point) or trash (0 points). Participants were instructed to earn the most points possible by learning (through trial-and-error) and choosing the best choice option. Participants were also instructed that the best option could change during the task, but were not informed about how often or when this would occur (see Supplementary Materials A for the complete instructions). Prior to beginning the experiment, participants had to correctly answer four comprehension questions about the instructions. Failing to correctly answer all items forced the participant to start the instructions over.

@@

\subparagraph{Clinical cutoffs.} We compared the proportion of participants in our sample reaching the threshold for clinical symptomology before and after applying exclusions. For the GAD-7, previous research has suggested a clinical cutoff score of 10 or higher \cite{lowe2008validation, hinz2017psychometric}. Though the 7-up/7-down scales do not have firmly established clinical cutoffs recent work has suggested a cutoff score of 12 or higher \cite{youngstrom2020evaluating}, which we use here. Finally, the original authors of the SHAPS scale recommended as a cutoff a score of 3 or more when the items are binarized (1, `Strongly disagree' or `Disagree'; 0, `Strongly agree' or `Agree'). We use this scoring approach in Table \ref{tab:tab02}.

@@

where $y_i$ is a participant's response to item $i$, $y_i$ is a participant's average score across all  items, and $k$ is the total number of items for a self-report measure. A composite ISD measure was estimated per participant by summing across each of the seven self-report scales. Larger ISD values indicate lower response consistency.

@@

where $r_t$ is the observed reward on trial $t$. In the risk-sensitive temporal difference learning model, there are separate learning rates for positive and negative prediction errors, such that positive and negative prediction errors have asymmetric effects on learning. For example, the effect of negative prediction errors on learned values is larger than that of positive errors if $\eta_p+ < \eta_n$, and vice versa if $\eta_p > \eta_n$.

@@

\textbf{SZ:} Conceptualization (equal); Software development (lead); Data collection - online (lead); Formal analysis (lead); Writing - Original Draft (lead); Writing - Review \& Editing (supporting); Visualization (lead). \textbf{JS:} Software development (supporting); Data collection - clinical (lead); Writing - Review \& Editing (supporting). \textbf{YN:} Writing - Review \& Editing (equal); Funding acquisition. \textbf{DB:} Conceptualization (equal); Software development (supporting); Data collection - online (supporting); Formal analysis (supporting); Writing - Review \& Editing (equal); Visualization (supporting).
